Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of neutron star - black hole binary observations

### 2016-09-07 Paper/paper.tex
progress on discussion

The pioneering terrestrial observation of gravitational waves by Advanced LIGO
harbingers the dawn of an era of gravitational-wave astronomy where observations
would finally drive scientific discovery~\cite{Abbott:2016blz}. As confirmed by
the first observations~\cite{Abbott:2016blz,Abbott:2016nmj,Abbott:2016nhf},
stellar-mass compact binary mergers emit GWs right in the sensitive frequency
band of the LIGO observatories, and are therefore their primary targets.
Neutron star black hole binaries form a physically distinct sub-class of
compact binaries, that we expect to detect the first of in the upcoming
observing runs~\cite{Abbott:2016ymx}. And when the detectors reach their design
sensitivity, we can expect to detect NSBHs at a healthy rate of $0.2-300$
mergers a year~\cite{Abadie:2010cf}.

\prayush{\bf Why NSBHs? 1. GRB, 2. tidal measurements.}

\prayush{\bf what we do in this paper - brief summary}

\prayush{\bf Sec 3 summary}
% % On the known NS and BH properties
% The number of neutron stars observed using conventional astronomical methods has
% grown rapidly in the recent past, both in isolated and two-body
% systems~\cite{Demorest:2010bx,Lyne:2004cj,2013Sci...340..448A,atnfcatalog,
% mcgillmagnetarcatalog,stellarcollapsemass}. Their masses span the range $1.2-2M_\odot$
% with an average of $\sim 1.35M_\odot$~\cite{Lattimer:2004sa,stellarcollapsemass}.
% On the other hand, NS spins have been observed to have magnitudes below
% $|\vec{S}_\mathrm{NS}|/\mns^2 < 0.01$~\cite{Miller:2014aaa}.
% % 
% On the other hand, indirect observations of stellar-mass BHs place their
% masses between $5-35M_\odot$, with their spin angular momenta 
% $|\vec{\chi}_\mathrm{BH}|$ ranging from small to nearly extremal (Kerr) values
% (see, e.g., Refs.~\cite{McClintockEtAl:2006,Miller:2009cw,Gou:2014una} for 
% examples of nearly extremal estimates of BH spins, Refs.~\cite{McClintock:2013vwa,
% Reynolds:2013qqa} for recent reviews of astrophysical BH spin measurements,
% and Figure 5 of Ref.~\cite{Miller:2014aaa} for a comparison of NS and BH spins).

\prayush{
We find that non-inclusion of tidal effects leads to systematic biases in
the measurement of non-tidal parameters. Consider first the binary chirp 
mass $\mchirp:=M\eta^{3/5}$ where $M$ is the binary total mass, and 
$\eta:=\mbh\mns/M^2$ is the dimensionless mass-ratio. We find that for
realistic (moderate) SNRs, i.e. $\rho\lesssim 30$, not including tidal
effects introduces a barely measurable bias that remains below $50\%$
of the statistical uncertainty of the $\mchirp$ measurement itself.
For higher signal strengths, we find that the underlying statistical
uncertainty decreases enough for the systematic bias to become comparable
to it for $\rho=50-70$ and higher. This is expected because chirp mass is fairly
accurately determined by the inspiral, where tidal effects are weak.
Next, we consider the mass-ratio $\eta$. We find that only if the NS is quite
deformable (with $\lambdans=2000$), {\it and} has a companion BH whose mass
is $\mbh\lesssim 4.5M_\odot$ (i.e. in the astrophysical ``mass-gap''~\cite{
Bailyn:1997xt,Kalogera:1996ci,Kreidberg:2012,Littenberg:2015tpa}),
will the systematic bias in $\eta$ measurement become comparable to its
statistical uncertainties at SNRs $\rho\leq 30$. For SNRs $\rho\gtrsim30$,
however, not including tidal effects is very likely to {\it significantly} bias
the mass-ratio estimate for NSBH binaries. Finally, we consider the spin on the
black hole $\chibh$. We find that $\chibh$ measurements are affected in a similar
way to $\eta$. By not including tidal effects, we compromise our measurement of
$\chibh$ significantly only if signal $\rho\gtrsim 30$. 
}

\prayush{
Overall, we learn an important thing. Bayesian parameter estimation algorithms
aimed at identifying NSBH binaries with low latency in order to alert 
electromagnetic (EM) observers in time to establish a possible coincident 
observation of a GRB~\cite{2012A&A...541A.155A,Singer:2014qca,Singer:2015ema,
Pankow:2015cra} can use BBH templates to look for NSBH systems with aLIGO.
This is so because the primary requirement of the EM follow-up effort is rapid
classification of a binary as an NSBH system, which can be achieved just as 
easily with BBH templates, purely on the basis of the smaller component's mass.
Another conclusion we draw is that detection searches are unaffected by the
choice of ignoring tidal effects in matched-filtering templates.
}

\prayush{\bf Sec 4 summary}

@@

\prayush{%
We find that when the BH is sufficiently spinning in the prograde sense,
with $\chibh\gtrsim +0.7$, even at a relatively possible signal strengths ($\rho=20$),
we can begin to constraint $\lambdans$ at the level of $\pm 75\%$ errors with a $90\%$
credible level. If the signal is stronger, say $\rho=30$, we can contraint
$\lambdans$ better with only $\pm 50\%$ errors, for binaries with large BH spins.
This trend continues as we vary the SNR from $\rho=30-50$ upwards. In other words,
with a single but moderately loud NSBH signal observation, Advanced LIGO will begin
to put constraints on NS matter properties. These constraints can consequently
be used to assess the likelihood of different equations of state for nuclear
matter, and possibly narrow the range they span.
}

>>>>

The pioneering terrestrial observation of gravitational waves by Advanced LIGO
harbingers the dawn of an era of gravitational-wave astronomy where observations
would finally drive scientific discovery~\cite{Abbott:2016blz}. As confirmed by
the first observations~\cite{Abbott:2016blz,Abbott:2016nmj,Abbott:2016nhf},
stellar-mass compact binary mergers emit GWs right in the sensitive frequency
band of the LIGO observatories, and are their primary targets.
Neutron star black hole binaries form a physically distinct sub-class of
compact binaries. We expect to detect the first of them in the upcoming
observing runs~\cite{Abbott:2016ymx}, and subsequently at a healthy rate of
$0.2-300$ mergers a year when aLIGO detectors reach design
sensitivity~\cite{Abadie:2010cf}.

NSBH binaries are interesting for various reasons. Unlike BBHs, the presence of
matter allows for richer phenomena to occur alongside the strong-field
gravitational dynamics. The quadrupolar moment of the NS changes during the
course of inspiral, which increases the inspiral rate of the binary and alters the
form of the emitted gravitational waves. Close to merger, under restricted but
plausible conditions, the neutron star is disrupted by the tidal field of its 
companion black hole and forms an accretion disk around it. This disruption
reduces the quadrupolar moment of the system, and decreases the amplitude of
the emitted GWs from the time of disruption through to the end of ringdown.
Both of these phenomena are discernible in their gravitational-wave signatures
alone. In addition, if the neutron star matter is magnetized, the magnetic
winding above the remnant black hole poles can build up magnetic fields
sufficiently to power short gamma-ray bursts (SGRB)~\cite{}.
Therefore a coincident observation of gravitational waves from an NSBH merger
and a SGRB can potentially confirm the hypothesis that the former is a
progenitor of the latter~\cite{}.

In this paper we study the observability of tidal signatures in the
gravitational-wave spectrum of NSBH binaries. More specifically, we investigate
three questions. First, what is the effect of not including tidal effects in 
templates while characterizing NSBH signals? Second, if we do include tidal 
effects, how well can we measure the tidal deformability of the NS
(parameterized by $\lambdans$) from individual NSBH signals? And third, as we
observe more and more signals, how does our knowledge of $\lambdans$ improve?
In the following, we summarize our main findings.

First, we study the effects of not including tidal terms in our search
templates while characterizing NSBH signals. We expect that the waveform
template that best fits the signal would compensate for the reduced number of
degrees of freedom in the template model by moving away from the true
parameters of the binary. This should result in a {\it systematic} bias in 
the recovered values of non-tidal source parameters, such as its masses 
and spins. In order to quantify it, we inject tidal signals into zero noise,
and perform a Bayesian parameter estimation analysis on them using templates
{\it without} tidal terms.
% 
We use the LEA+ model to produce tidal waveforms that incorporate the effect
of NS distortion during inspiral, and of its disruption close to merger. Our
injected signals sample the region of NSBH parameter space where NS disruption
prior to binary merger is likely {\it and} can be modeled using LEA+. Their
parameters are given by combinations of $q=\mbh/\mns=\{2,3,4,5\},
\chibh=\{-0.5, 0, 0.5, 0.75\}$ and $\lambdans=\{500,1000,1500,2000\}$.
Other parameters, such as source location and orientation, that factor out of
$h(t)$ as amplitude scaling are co-sampled by varying $\rho=\{20,30,50,70\}$.

At low to moderate SNRs ($\rho\lesssim 30$), we find that using BBH templates
does not significantly hamper our estimation of non-tidal parameters for NSBH
signals. In the worst case, when the BH mass is within the astrophysical mass-gap
and its spin is positive aligned, the systematic biases in $\eta$ and $\chibh$
measurements do become somewhat comparable to statistical errors (ratio
$\sim 0.5-0.8$) under very restrictive conditions~\footnote{requiring a
companion BH with mass $\mbh\lesssim 4.5M_\odot$ (i.e. in the astrophysical
``mass-gap''~\cite{Bailyn:1997xt,Kalogera:1996ci,Kreidberg:2012,
Littenberg:2015tpa}), and the
hardest NS EoS considered (with $\lambdans\simeq 2000$).}, but never exceed them.
At high SNRs ($\rho\gtrsim 50$), systematic biases in $\mchirp$ become larger
than the statistical uncertainties. For $\eta$ and $\chibh$ the difference
is more drastic with the systematics reaching up to $4\times$ the statistical
errors. We therefore conclude that $\rho\simeq 30-50$ is loud enough to
motivate the use of tidal templates for even the estimation of non-tidal
parameters from NSBH signals.
% 
We also conclude that low-latency parameter estimation algorithms, designed to
classify GW signals into electromagnetically active (NSBH and NSNS) and
inactive (BBH) sources, can use BBH templates to trigger GRB 
alerts~\cite{2012A&A...541A.155A,Singer:2014qca,Singer:2015ema,Pankow:2015cra,
Abbott:2016wya,Abbott:2016gcq} for NSBH signals with low to moderate SNRs
($\rho\lesssim 30$).
This is so because the primary requirement of identifying NS-X binaries (X =
\{NS, BH\}) can be achieved just as easily with BBH templates, on the basis of
the smaller component's mass\footnote{The smaller component mass is unlikely
to be significantly biased by missing tidal effects in filter templates below
$\rho\simeq 30$, as we show above.}.
% 
We also speculate that NSBH detection searches are unlikely to be
affected by the choice of ignoring tidal effects in matched-filtering
templates, if these effects are too subtle to manifest in parameter estimation
below $\rho\simeq 30$.

% 
% % On the known NS and BH properties
% The number of neutron stars observed using conventional astronomical methods has
% grown rapidly in the recent past, both in isolated and two-body
% systems~\cite{Demorest:2010bx,Lyne:2004cj,2013Sci...340..448A,atnfcatalog,
% mcgillmagnetarcatalog,stellarcollapsemass}. Their masses span the range $1.2-2M_\odot$
% with an average of $\sim 1.35M_\odot$~\cite{Lattimer:2004sa,stellarcollapsemass}.
% On the other hand, NS spins have been observed to have magnitudes below
% $|\vec{S}_\mathrm{NS}|/\mns^2 < 0.01$~\cite{Miller:2014aaa}.
% % 
% On the other hand, indirect observations of stellar-mass BHs place their
% masses between $5-35M_\odot$, with their spin angular momenta 
% $|\vec{\chi}_\mathrm{BH}|$ ranging from small to nearly extremal (Kerr) values
% (see, e.g., Refs.~\cite{McClintockEtAl:2006,Miller:2009cw,Gou:2014una} for 
% examples of nearly extremal estimates of BH spins, Refs.~\cite{McClintock:2013vwa,
% Reynolds:2013qqa} for recent reviews of astrophysical BH spin measurements,
% and Figure 5 of Ref.~\cite{Miller:2014aaa} for a comparison of NS and BH spins).

Second, we turn the question around to ask: can we measure the tidal effects if
our template models did account for them? Tidal effects in our waveform model
are parameterized using a single deformability parameter 
$\lambdans\propto (R/M)_\mathrm{NS}^5$. In order to quantify the 
measurability of $\lambdans$, we inject the same tidal signals as before, and
this time perform a Bayesian analysis on them using {\it tidal} templates. 
The results are detailed in Sec.~\ref{s1:PEwithNS}.
% 
At low SNRs ($\rho\simeq 20$), we find that the best we can do is to constrain
$\lambdans$ within $\pm 75\%$ of its true value at $90\%$ credible level. This
too only if the BH is spinning sufficiently rapidly, with $\chibh\gtrsim +0.7$,
and the NS has $\lambdans\gtrsim 1000$. At moderate SNRs ($\rho\simeq 30$), we
can constrain $\lambdans$ a little better, i.e. within $\pm 50\%$ of its true
value. This level of accuracy, however, again requires that BH spin
$\chibh\gtrsim+0.7$ and $\lambdans\gtrsim 1000$. Binaries with smaller BH spins
and/or softer NS EoSs will furnish worse than $\pm 75\%-\pm 100\%$ errors for
$\lambdans$. This trend continues as we increase the SNR from $\rho=30-50$. It
is not before we reach an SNRs as high as $\rho\simeq 70$ that we can shrink
$\lambdans$ errors substantially with a single observation (i.e. within
$\pm 25\%$ of its true value).
% 
In summary, we find that with a single but moderately loud NSBH signal,
Advanced LIGO can begin to put a factor of $1-2\times$ constraints on NS tidal
deformability parameter. These constraints can subsequently be used to assess
the likelihood of various candidate equations of state for nuclear matter, and
possibly to narrow the range they span.

@@
